---
title: "Programming with the HCUP SID"
institute: "Using Administrative Data for<br>Clinical and Health Services Research"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  message = FALSE
)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
xaringanthemer::style_solarized_dark()
```

```{r knit, eval = FALSE, include = FALSE}
rmarkdown::render("./R-lecture-01-intro-programming-SID.Rmd")
```


# SID Data

- Florida SID Q4 2015&ndash;2019

  - Transition to ICD-10 in October 2015
  
--
  
- Records are all admissions from acute care community hospitals

  - Year (`YEAR`) and quarter (`DQTR`) refer to discharge
  
  - All-payer
  
--

- Florida hospitals, but patients not necessarily Florida residents

--

- Patients may have multiple records, i.e., multiple admissions

  - Duplicates are rare but do occur
  
    - Most common are interim discharge records, with same patient and
      admit date but different information
      

---


# Data Files

- **Core** files contain most admission-related information

--

- **Charges** files record services provided that aren't captured by
  diagnosis/procedure codes
  
  - Correspond to revenue center codes
  
--
  
- **AHA Linkage** files associate hospital identifiers with those in
  the annual American Hospital Association survey databases
  
--

- HCUP **Cost-to-Charge Ratio** for Inpatient Files contain multipliers for converting from *charges* to *cost estimates*

  - `HTYPE` codes hospital ownership, size, and urban/rural status

  
---

  
# Identifiers

- Each record in Core and Charges has unique `KEY`

--

- Each patient has unique `VisitLink`

  - Upon entering data, random number `DaysToEvent`
    assigned representing admission date
    
	- Indicates timing of subsequent events, e.g., admission 120 days later will have `DaysToEvent + 120`
    
  - `VisitLink` and `DaysToEvent` are retained across years
  
  - Quarter `DQTR` is as specific as we can be regarding calendar date

--

- In Core files, hospital is identified by `DSHOSPID`

--

  - AHA Linkage files can be used to get `HOSPID` (for cost-to-charge ratio files) and `AHAID` (for AHA survey files)


---


### Core File Variables

- Over 200 variables

--

- Admission type (`ATYPE`) and disposition at discharge (`DISPUNIFORM`)
  - Hour of admission (`AHOUR`) and indicator for weekend admission (`AWEEKEND`)
  
--
  
- Expected primary payer (`PAY1`)

--

- Demographics:
  - `AGE`
  - `FEMALE`
  - `RACE`, `HISPANIC`
  - `PL_CBSA`, `PL_NCHS`, `PL_RUCC`, `PL_UIC`, `PL_UR_CAT4`
  - County and zip code
  
--

- Socioeconomic status:
  - `Homeless`
  - `MEDINCSTQ`, `ZIPINC_QRTL`


---


# Diagnosis Codes

- *International Classification of Diseases, Tenth Revision, Clinical Modification* (ICD-10-CM) codes
  - Recorded in `I10_DX`*n* fields, between 31 and 34 per year
  
--

### E Codes

- External cause of morbidity codes (ICD-10: V00&ndash;Y99)
  - Recorded in `I10_ECAUSE`*n* fields prior to 2017
  - Placed with other codes in `I10_DX`*n* fields starting in 2017
  
--
  
### Present on Admission (POA) Codes

- `DXPOA`*n* and `E_POA`*n* fields
  - See documentation for meaning of different values


---


# Procedure Codes

- *International Classification of Diseases, Tenth Revision, Procedure Coding System* (ICD-10-PCS) codes
  - Recorded in `I10_PR`*n* fields, 31 per year
  - No CPT/HCPCS codes
  
--

### Timing Procedures

- `PRDAY`*n* fields record when procedures took place with respect to admission
  - 0 indicates day of admission, 1 indicates day after, ...
  - Negative values indicate days prior to admission


---


# Resources

- SID database documentation

  - https://www.hcup-us.ahrq.gov/db/state/siddbdocumentation.jsp
  
- SID data elements by year (i.e., data dictionary)

  - https://www.hcup-us.ahrq.gov/db/state/siddist/siddist_ddeavailbyyear.jsp

- SID cost-to-charge ratio for inpatient files documentation

  - https://www.hcup-us.ahrq.gov/db/ccr/ip-ccr/ip-ccr.jsp

- List of variables in 1% SID data set with SAS labels

  - admin_course_data/resources/variables_1p_sample.xlsx
  
- List of variables in full SID data set with SAS labels

  - admin_course_data/resources/variables_SID.xlsx
  - admin_course_data/resources/variables_comparison_by_year_SID.xlsx
