---
title: "Merging data using dplyr"
institute: "Using Administrative Data for<br>Clinical and Health Services Research"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r knit, eval = FALSE, include = FALSE}
rmarkdown::render("./R-lecture-03-merging-data-with-dplyr.Rmd")
```

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  message = FALSE
)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
style_solarized_dark()
```

```{r demo-setup, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

options(width = 70)
payer_info <- tribble(
  ~VisitLink, ~PAY1,
  46571, 1,
  93576, 3,
  69250, 3
)
charge_info <- tribble(
  ~VisitLink, ~TOTCHG,
  46571, 28004,
  93576, 7140,
  73038, 82450
)
payer_info2 <- tribble(
  ~VisitLink, ~PAY1, ~AGE,
  46571, 1, 25,
  93576, 3, 31,
  69250, 3, 44
)
charge_info2 <- tribble(
  ~VisitLink, ~TOTCHG, ~AGE,
  46571, 28004, 25,
  93576, 7140, 31,
  73038, 82450, 76
)
```


# Overview

- Merging in the SID

- Key variables

- Examples


---


# Merging in the SID

- Most variables of interest available in the CORE files

- Exceptions:

  - Detailed charges
  
  - Hospital characteristics
  
  - External data (e.g., from Census Bureau)

--

### Keys

- Variables used to merge data sets are referred to as *keys*

--

- Merging requires understanding of how data are organized and what variable(s) represent keys

- `KEY` in the SID makes this easy for discharge *records*


---


# Examples

```{r}
payer_info
charge_info
```


---


# Inner Join

- Combines observations with matches in the key variable(s)

  - Unmatched records are dropped

--

```{r}
payer_and_charge_info <- inner_join(
  payer_info, charge_info, by = "VisitLink")
payer_and_charge_info
```

--

- Or using `%>%`,

```{r}
payer_and_charge_info <- payer_info %>%
  inner_join(charge_info, by = "VisitLink")
```


---


# Left Join

- Combines observations with matches in the key variable(s)

  - Unmatched records in 'left' data set are retained, those in 'right' data set are dropped

--

```{r}
payer_and_maybe_charge_info <- payer_info %>%
  left_join(charge_info, by = "VisitLink")
payer_and_maybe_charge_info
```

--

- `right_join` works in an analogous way


---


# Full join

- Combines observations with matches in the key variable(s)

  - Unmatched records in both data sets are retained

--

```{r}
payer_or_charge_info <- payer_info %>%
  full_join(charge_info, by = "VisitLink")
payer_or_charge_info
```


---


# Anti Join

- *Removes* observations with matches in the key variable(s)

--

```{r}
missing_payer_info <- payer_or_charge_info %>%
  filter(is.na(PAY1))
missing_payer_info
```

--

```{r}
charge_info_but_only_if_payer_info <- charge_info %>%
  anti_join(missing_payer_info, by = "VisitLink")
charge_info_but_only_if_payer_info
```


---


# Merging Tip

- Make sure the only shared variables are the key variable(s)

```{r}
payer_info2
charge_info2
```


---


# Merging Tip

- Make sure the only shared variables are the key variable(s)

```{r}
payer_and_charge_info_with_duplicate_age <- payer_info2 %>%
  inner_join(charge_info2, by = "VisitLink")
payer_and_charge_info_with_duplicate_age
```


---


# Merging Tip

- Make sure the only shared variables are the key variable(s)

```{r}
payer_and_charge_info_without_duplicate_age <- payer_info2 %>%
  select(!AGE) %>%
  inner_join(charge_info2, by = "VisitLink")
payer_and_charge_info_without_duplicate_age
```


---


# Merging Tip

- Alternatively, add shared variables to the key

```{r}
payer_and_charge_info_without_duplicate_age <- payer_info2 %>%
  inner_join(charge_info2, by = c("VisitLink", "AGE"))
payer_and_charge_info_without_duplicate_age
```
