---
title: "Identifying Diagnosis and Procedures Codes in R"
institute: "Using Administrative Data for Clinical and Health Services Research"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r knit, eval = FALSE, include = FALSE}
rmarkdown::render("./R-lecture-04-strings-dx-pr-codes.Rmd")
```

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  message = FALSE
)

library(tidyverse)

core1p <- read_rds(
  paste0(
    "//storage1.ris.wustl.edu/colditzg/Active/",
    "admin_course_data/fl_sidc_core_1pSample.rds")
)
```

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)
style_solarized_light()
```


# Overview

- ICD-10-CM/PCS from a programming perspective

- Approaches for the SID data


---


# ICD-10-CM

- Over 68,000 codes at time of transition from ICD-9-CM (just over 14,000 codes)

- 3&ndash;7 characters
  - First character is letter
  - Second character is number
  - 3&ndash;7 are letters or numbers
  
- Decimal point between third and fourth characters
  - But not in HCUP fields

--

- Inexact mapping between ICD-9-CM and ICD-10-CM
  - One ICD-9 $\to$ many ICD-10
  - Many ICD-9 $\to$ one ICD-10


---


# ICD-10-PCS

- Over 72,000 codes at time of transition from ICD-9-CM (less than 4,000 codes)

- Always 7 characters
  - Z placeholder for when a code contains fewer than 7 characters
  
- All characters can be letters or numbers
  - First character is section of code book
  - Second character is body system
  - Third character is root operation
  - Fourth character is body part
  - Fifth character is approach
  - Sixth character is device
  - Seventh character is qualifier

- Similar problems with inexact matching between coding systems


---


# Identifying codes in the SID

- Goal is to determine whether a particular diagnosis or procedure was coded during an admission

- Do any of the codes recorded in `I10_DX`n / `I10_ECAUSE`n or `I10_PR`n match any of the codes for our diagnosis or procedure of interest?


---


# Identifying codes in the SID

- Suppose we're interested in acute myocardial infarction

- I21.0&ndash;I21.4, I22.0&ndash;I22.2, I22.8, I22.9

```{r acutemi-codes}
dx10_acutemi <- c(
  "I2101", "I2102", "I2109", "I2111", "I2119", "I2121", "I2129",
  "I213", "I214", "I220", "I221", "I222", "I228", "I229"
)
```


---


# Identifying codes in the SID

- Start with single admission, single code

```{r}
my_data <- data.frame(KEY = 1, DX1 = "I220")
```

- One way to compare values is with `==`

```{r eval = FALSE}
my_data %>%
  mutate(
    dx_acutemi = ifelse(DX1 == "I2101" | DX1 == "I2102" | <...>, 1, 0)
  )
```

--

- More efficient way is with `%in%`

```{r}
my_data %>%
  mutate(
    dx_acutemi = ifelse(DX1 %in% dx10_acutemi, 1, 0)
  )
```


---


# Identifying codes in the SID

- What if we have multiple admissions each with a single code?

```{r}
my_data <- data.frame(
  KEY = 1:2,
  DX1 = c("I220", "E0800")
)

my_data %>%
  mutate(dx_acutemi = ifelse(DX1 %in% dx10_acutemi, 1, 0))
```


---


# Identifying codes in the SID

- We have multiple columns of diagnosis and procedure codes for each admission

```{r}
my_data <- data.frame(
  KEY = 1:3,
  DX1 = c("I220", "I209", "J80"),
  DX2 = c("E0800", "I220", "O9921")
)

my_data %>%
  mutate(
    dx_acutemi = ifelse(
      DX1 %in% dx10_acutemi | DX2 %in% dx10_acutemi, 1, 0)
  )
my_data
```


---


# Identifying codes in the SID

- So real problem is that we have LOTS of columns with diagnosis and procedure codes and likely many diagnoses and procedures that we'd like to identify

```{r eval = FALSE}
core1p2 <- core1p %>%
  mutate(
    dx_acutemi = case_when(
      I10_DX_Admitting %in% dx10_acutemi ~ 1,
      I10_DX1 %in% dx10_acutemi ~ 1,
      I10_DX2 %in% dx10_acutemi ~ 1,
#      ...etc...
      I10_DX33 %in% dx10_acutemi ~ 1,
      I10_DX34 %in% dx10_acutemi ~ 1,
      I10_ECAUSE1 %in% dx10_acutemi ~ 1,
      I10_ECAUSE2 %in% dx10_acutemi ~ 1,
      I10_ECAUSE3 %in% dx10_acutemi ~ 1,
      I10_ECAUSE4 %in% dx10_acutemi ~ 1,
      I10_ECAUSE5 %in% dx10_acutemi ~ 1,
      I10_ECAUSE6 %in% dx10_acutemi ~ 1,
      TRUE ~ 0
    )
  )
```


---


# Identifying codes in the SID

- Things are much easier if we're only concerned with 'primary' (i.e., in `I10_DX1` only) diagnoses of our condition of interest

```{r eval = TRUE}
core1p2 <- core1p %>%
  mutate(
    dx1_acutemi = ifelse(I10_DX1 %in% dx10_acutemi, 1, 0)
  )
table(core1p2$dx1_acutemi, useNA = "ifany")
```


---


# Identifying codes in the SID

- Identifying procedures works in a similar way

``` r
core1p2 <- core1p %>%
  mutate(
    pr_cabg = case_when(
      I10_PR1 %in% pr10_cabg ~ 1,
      I10_PR2 %in% pr10_cabg ~ 1,
#     ...etc...
      I10_PR30 %in% pr10_cabg ~ 1,
      I10_PR31 %in% pr10_cabg ~ 1,
      TRUE ~ 0
    )
  )
```

- No special significance to the code in `I10_PR1`


---


# Options for Programming

- Lengthy, but relatively easy to copy and modify

  - Create a vector of codes like `dx10_acutemi`
  
  - Must list the codes that actually appear on records
  
``` r
# Bad
dx10_major_depress <- "F33"
# Good
dx10_major_depress <- c(
  "F330", "F331" "F332", "F333", "F3340", "F3341", "F3342",
  "F338", "F339"
)
```

---


# Options for Programming

- Lengthy, but relatively easy to copy and modify

  - Create a vector of codes like `dx10_acutemi`
  
  - Must list the codes that actually appear on records

``` r
core1p_with_major_depress <- core1p %>%
  mutate(
    dx_major_depress = case_when(
      I10_DX_Admitting %in% dx10_major_depress ~ 1,
      I10_DX1 %in% dx10_major_depress ~ 1,
#     ...etc...
      )
  )
```

---


# Options for Programming

- Functions defined that uses the same code

  - Saved in **admin_course_data/code2023/**

--

- Read definitions from external R code file with `source`

``` r
source(
  paste0(
    "//storage1.ris.wustl.edu/colditzg/", # <- different for Mac users
    "Active/admin_course_data/code2023/coder.R"
  )
)
```


---


# Options for Programming

- All functions use the same first three arguments

  1. Name of input data set
  
  2. Name of variable with code list
  
  3. Name of new variable that will be made
  
- Return the input data set with the new column added

- Work similar to `mutate`, so each named `mutate_*`

---


# Options for Programming

- `mutate_flag_dx` for creating 0/1 variables for diagnoses

``` r
core1p_with_acutemi <- mutate_flag_dx(core1p, dx10_acutemi, dx_acutemi)
```

--

- is equivalent to

``` r
core1p_with_acutemi <- mutate(
  core1p,
  dx_acutemi = case_when(
    I10_DX_Admitting %in% dx10_acutemi ~ 1,
    I10_DX1 %in% dx10_acutemi ~ 1,
    I10_DX2 %in% dx10_acutemi ~ 1,
#   ...etc...
    I10_ECAUSE5 %in% dx10_acutemi ~ 1,
    I10_ECAUSE6 %in% dx10_acutemi ~ 1,
    TRUE ~ 0
  )
)
```


---


# Options for Programming

- `mutate_flag_dx` for creating 0/1 variables for diagnoses

``` r
core1p_with_acutemi <- mutate_flag_dx(core1p, dx10_acutemi, dx_acutemi)
```

- Can also be written using `%>%`

``` r
core1p_with_acutemi <- core1p %>%
  mutate_flag_dx(dx10_acutemi, dx_acutemi)
```

--

- Can define variables for multiple diagnoses/procedures in same block of code

``` r
core1p2 <- core1p %>%
  mutate_flag_dx(dx10_acutemi, dx_acutemi) %>%
  mutate_flag_dx(dx10_istroke, dx_istroke)
```


---


# Options for Programming

- `mutate_flag_dx1` for creating 0/1 variables for primary diagnoses

``` r
core1p_with_dx1_acutemi <- core1p %>%
  mutate_flag_dx1(dx10_acutemi, dx1_acutemi)
```

--

- `mutate_flag_pr` for creating 0/1 variables for procedures

``` r
core1p_with_cabg <- core1p %>%
  mutate_flag_pr(pr10_cabg, pr_cabg)
```

- Other functions will be shown in future lectures


---


# References

- HCUP introduction to ICD-10 coding systems
https://www.hcup-us.ahrq.gov/datainnovations/BriefIntrotoICD-10Codes041117.pdf

- Centers for Medicare and Medicaid Services General Equivalence Mappings (for translating ICD-9 to ICD-10)
https://www.cms.gov/Medicare/Coding/ICD10/2018-ICD-10-CM-and-GEMs
